Effects of prenatal CORT and temperature on the response to a stressor

### 2025-07-10 ms.qmd
updates

Animals constantly face events that can disrupt their homeostasis. These disruptions can be acute, such as a sudden predator attack, or long-term, like prolonged droughts or habitat degradation [@sapolsky_how_2000; @mcewen_concept_2003]. Regardless of the nature of the challenge, organisms must adjust to allostasis to maintain their physiological and behavioural integrity [@wingfield2002endocrine]. In vertebrates, the ability to cope with disruptions in homeostasis is carried out by the hypothalamic-pituitary-adrenal/interrenal (HPA/HPI) axis [@sapolsky_how_2000]. This system elicits a physiological response, also known as the stress response, mediated by glucocorticoids (GCs), a group of hormones that facilitate the reallocation of energetic resources to restore homeostasis [@sapolsky_how_2000; @mcewen_concept_2003]. For example, when an animal faces a predatory attack, GCs mobilise energy to support increased locomotor activity, vigilance, or attention to avoid the threat [@trompeter2011invader]. In the short term, GCs help organisms cope with acute challenges. However, long-term exposure to GCs can carry significant metabolic costs [@picard_mitochondria_2014]. Prolonged elevations in GCs can lead to immune suppression, impaired growth, or permanent neural alterations [@mcewen2017neurobiological]. Furthermore, reacting to stressors can reduce the time and energy available for essential activities such as foraging, thermoregulation, or social interactions [@belliure2004behavioral; @martin2024blind]. As such, animals are predicted to evolve mechanisms to appropriately respond to acute stressors while avoiding the costs in the long term. For example, by regulating their responses to homotypic (repeated and similar) stressors, particularly those that are predictable and non-lethal [@grissom2009habituation]. Habituation - the reduction of physiological responses elicited by exposure to a homotypic stressor [@pfister1979glucocorticosterone] - can be crucial to avoid the long-term costs of chronic stress without compromising the response to acute stressors.

Importantly, the process of physiological habituation shares key functional characteristics with the cognitive process of habituation - a form of non-associative learning where individuals reduce their responsiveness to a repeated stimulus [@thompson1966habituation]. In both cases, organisms are repeatedly exposed to a stimulus/homotypic stressor and progressively decrease their response towards it [@grissom2009habituation]. In fact, both behavioural and physiological habituation often co-occur. For example, rhesus macaques (_Macaca mulatta_) showed a decline in blood cortisol concentrations across consecutive days of restraint together with decreased vocalizations and behavioral agitation [@ruys2004behavioral]. How both types of habituation interplay with each other depends on the nature of the stressor. Physiological habituation does not occur when animals are faced with systemic stressors like hypoglycemia or ether exposure, whereas habituation to processive stressors — those involving primarily psychological qualities, such as restraint or simulated predation - is frequent [@grissom2009habituation]. However, physiological habituation alone is insufficient to explain the behavioural patterns observed when animals are repeatedly exposed to a homotypic processive stressor [@jaferi2006corticosterone; @grissom2009habituation]. Adrenalectomised rats - incapable of physiological habituation - still exhibit cognitive habituation [@jaferi2006corticosterone], suggesting a degree of independence between the two processes. While the HPA/HPI axis is paramount for physiological habituation, cognitive habituation also involves neural mechanisms related to sensory processing, learning, and behavioural control, such as the limbic system, the superior colliculus, or the optic tectum [@grissom2009habituation; @dutta2014saliency]. Habituation to processive stressors is therefore likely to involve complex interactions between physiological and cognitive processes.

Crucially, the effectiveness of either system depends on how well the underlying neural structures develop and integrate. Many of these brain regions mature concurrently and can be highly plastic during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017]. Early-life conditions can shape the structure and function of both neural systems, with long-term consequences for how individuals respond to acute and chronic stressors [@vargas2016early; @van2017modulation; @zito2017early; @crino_under_2020; @crino2024eggs]. One of the key mediators of early-life effects is GC exposure. Elevations in GCs during development exert potent effects on HPA axis function, altering the response to stressors later in life [@costantini2011meta; @vargas2016early; @zito2017early; @crino2024eggs]. For example, zebra finches (_Taeniopygia guttata_) nutritionally stressed during early-life had higher levels of CORT than controls at post-hatching day 25, while periodic maternal separation increased basal corticosterone levels in rats (_Rattus norvegicus_) [@vargas2016early; @zito2017early]. In parallel, developmental GC exposure can impact brain regions involved in learning, memory, and behavioural flexibility, affecting cognitive processes relevant to cognitive habituation [@lemaire_prenatal_2000; @zhu_prenatal_2004; @costantini2011meta]. Therefore, early-life exposure to GCs can have long-lasting effects on the response to both acute and chronic stressors, potentially leading to maladaptive responses in adulthood.

@@

In this study, we investigate how prenatal CORT and incubation temperature shape the response to a stressor in two species of skinks: the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We focus on the effects of these early-life conditions on the behavioural responses to a simulated predatory attack. Specifically, we examine how these factors influence the acute response to a stressor (i.e., the first exposure) and the chronic response (i.e., repeated exposures over time). Finally, we assessed body mass change across the stress period as a potential indirect cost, while controlling for food intake. We predicted that prenatal CORT exposure would heighten acute stress reactivity and slow habituation, while warmer incubation temperatures would reduce acute responses and enhance habituation. Furthermore, we predicted higher temperatures would buffer the effects of prenatal CORT on stress responses, potentially leading to more flexible behavioural adjustment. Finally, we expected that less flexible stress responses would lead to higher decreases in mass.

@@

We used two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). These are oviparous, generalist, small (∼35-55 mm snout-vent length (SVL)) lizards that are sympatric in suburban areas throughout south-eastern Australia [@chapple_know_2011; @chapple_biology_2014]. However, _L. delicata_ has been reported as a successful invader in New Zealand and Hawaii, while there are no such reports for _L. guichenoti_ [@chapple_know_2011; @chapple_biology_2014]. Invassiveness does not seem to be related to differences in cognitive abilities [@bezzina2014does; @recio2025early; @recio2025early], but seems to be associated with differences in personality [@chapple_know_2011], which may be linked to differences in their stress response [@koolhaas1999coping].

@@

*Breeding colony* - Lizards came from a breeding colony established in the laboratory since 2019. This colony consisted of 270 adults of _L. delicata_ and 180 adults of _L. guichenoti_ housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with two males and four females per enclosure. Enclosures were provided with shelter, nonstick matting, and several small water dishes. The lizards were given water daily and were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. The crickets were dusted with calcium weekly and multivitamins and calcium biweekly. Room temperature was set to 22-24 ºC, but we also provided the enclosures with a heat cord and a heat lamp following a 12 h light:12 h dark cycle, keeping the warm side of the enclosures at 34 ºC.

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them at cold (23 ± 3 ºC) or hot (28 ± 3 ºC) conditions (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study, where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Mean incubation temperatures represent the lower and upper limits of the natural range of nest temperatures in _L. delicata_ [mean nesting temperatures = 27.4 ºC; @cheetham2011embryonic]. Nonetheless, the higher temperature treatment (28 ºC) was above the thermal optima estimated for _L. delicata_ [T~opt~ = 25 ºC; @pettersen2023maternal] and reached more stressful temperatures daily as fluctuations occurred (i.e., 31 ºC).

@@

To test the effects of prenatal conditions on the response to a stressor, we simulated a predatory attack - a standardized processive stressor [@thaker2009acute; @trompeter2011invader] - once daily over multiple days. Each trial began with one of the researchers (PR) removing the water dish from the enclosure, followed by the shelter. The lizard was then chased for 60 seconds using a soft paintbrush, simulating a predatory attack. Immediately after the chase, the shelter — but not the water dish — was returned to its original position, marking the start of the behavioural observation period. The lizard’s behaviour was then recorded for one hour. All videos were later analysed by MD, who was blinded to the lizards’ treatment. From each video, we recorded three behavioural variables: i) latency to move — the time spent immobile (frozen) immediately after the simulated attack; ii) latency to shelter — the time it took the lizard to hide under the shelter after resuming movement; and iii) emergence — whether the lizard emerged from the shelter during the 40 minutes following hiding (1 = emerged, 0 = did not emerge). A lizard was considered to have resumed movement if it moved continuously for at least five seconds, and to be under shelter if no limbs were visible. This procedure was repeated daily for eight consecutive days to assess the effects of mid-term exposure to a stressor. All trials were conducted between 1000–1400 h, when lizards are most active.

Both before and after the completion of the behavioural tests, lizard mass was measured to the nearest 0.001 g using a digital scale. Change in mass was calculated as the difference between the final minus the initial mass and then rescaled by adding the smallest mass change observed in the dataset, to avoid negative values. This rescaled change in mass was used as a proxy for the physiological cost of repeated stress exposure. To control for hunger levels and potential effects of food ingestion on mass change, we also recorded the number of crickets ingested by each lizard during the acclimation period. To do so, we placed a known number of crickets in each enclosure after each of the trials, and counted the number of crickets remaining after 24 hours. The total number of crickets ingested during the eight days of tests was then used as a covariate in the analyses of mass change. The experimental design is summarised in @fig-Methods B.

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Scheme of our experimental design. In panel A, the early environment manipulations leading to four experimental conditions. In panel B, the timeline of the postnatal stress-response assay. On Day 0, lizard mass was recorded. From Days 1–8, lizards were subjected daily to a simulated predator attack (chased with a soft paintbrush for 1 minute), followed by a 1-hour behavioural recording. Variables measured the latency to move, the latency to shelter, and the probability of emergening from the shelter in 40 minutes after the stressor. Final mass was measured on Day 9."

@@

We performed the analyses for each species separately. We fitted all models using the _brm_ package [@burkner2017brms], which fits Bayesian multilevel models with _Stan_ [@stan] with R version 4.4.0 [@R]. We ran four parallel MCMC chains of 10000 iterations for each model, with a warmup period of 4000 iterations.

We used as the response variables the latency to move, the latency to shelter, and the emergence probability, as well as the change in mass after the eight days of stress exposure. We fitted separate models for each response variable. For the three behavioural variables, we included the prenatal condition (Control-Cold, CORT-Cold, Control-Hot, CORT-Hot), the day of the trial, and their interaction as fixed effects. To facilitate interpretation of the model intercepts, trial day was transformed so that day 1 in the experiment corresponds to day 0 in the model (i.e., the intercept reflects lizard behaviour in response to the first exposure to the stressor). For the variable mass, we included the prenatal condition, the total number of crickets ingested during the eight days of stress exposure, and their interaction as predictors.

Lizard identity was included as a random factor in the behavioural models, but it was not included when mass was the response. For all models, we included the clutch as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple2015deliinvLHI]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

@@

@@

_Lampropholis delicata_: Incubation at cold temperatures decreased the latency to move after a simulated attack on the first day (Intercept (denoted hereafter as I)~Hot~ - I ~Cold~ = `r format_dec(exp(deli_move_contrasts$median_Temperature_int), 3)`; 95% Highest Posterior Density Intervals (95% HPDI) = [`r deli_move_contrasts$X2.5.HDI_Temperature_int`, `r deli_move_contrasts$X97.5.HDI_Temperature_int`]), while neither CORT or the interaction have any effect on latency to shelter or emergence (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_). We did not find any effects of prenatal temperature, CORT or their interaction on latency to shelter or emergence probability after an acute stressor (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_).

@@

```{r, fig-results_mass}
#| label: fig-results_mass
#| fig.cap: "Change in mass of Lampropholis delicata and Lampropholis guichenoti after eight days of repeated stress exposure." 
#
#
source(here("R", "func.R"))
#
## L delicata
deli_mass_plot <- plotting(sp = "deli", res = "mass", type = "int") +
  theme(legend.position = "none",
        legend.title = element_blank())
# Getting the legend
legend_plot_deli <- plotting(sp = "deli", res = "mass", type = "int") + 
  theme(legend.position = "right", legend.title = element_blank())
gtable_deli <- ggplotGrob(legend_plot_deli)
legend_deli <- gtable_deli$grobs[[which(sapply(gtable_deli$grobs, function(x) x$name) == "guide-box")]]
#
## L guichenoti
guich_mass_plot <- plotting(sp = "guich", res = "mass", type = "int") +
  theme(legend.position = "none",
        legend.title = element_blank())
# Getting the legend
legend_plot_guich <- plotting(sp = "guich", res = "mass", type = "int") + 
  theme(legend.position = "right", legend.title = element_blank())
gtable_guich <- ggplotGrob(legend_plot_guich)
legend_guich <- gtable_guich$grobs[[which(sapply(gtable_guich$grobs, function(x) x$name) == "guide-box")]]
#
mass_grid <- plot_grid(deli_mass_plot, NULL, guich_mass_plot, NULL,
                       ncol = 4, nrow = 1,
                       rel_widths = c(0.5, 0.5, 0.5, 0.45))
#
fig_mass <- ggdraw() +
  # Make background white
  theme(plot.background = element_rect(fill = "white", color = NA)) +
  # Add the final plot but with reduced size for the legend and "title"
  draw_plot(mass_grid, x = 0, y = 0, width = 1, height = 0.92) +
  # Insert species name
  annotate("text", x = 0.36, y = 0.99, label = "Lampropholis delicata", hjust = 1, vjust = 1, size = 5, fontface = "italic") +
  annotate("text", x = 0.9, y = 0.99, label = "Lampropholis guichenoti", hjust = 1, vjust = 1, size = 5, fontface = "italic") +
  # Insert title for each plot
  annotate("text", x = 0.023, y = 0.91, label = "A", hjust = 1, vjust = 1, size = 6, fontface = "bold") +
  annotate("text", x = 0.535, y = 0.91, label = "B", hjust = 1, vjust = 1, size = 6, fontface = "bold") +
  # Insert picture species
  draw_image(here("./Others/Deli.png"), x = 0.271, y = 0.65, width = 0.22, height = 0.22) +
  draw_image(here("./Others/Guich.png"), x = 0.78, y = 0.65, width = 0.22, height = 0.22) +
  # Insert legends
  draw_grob(legend_deli, x = 0.375, y = 0.275, width = 0.01, height = 0.01) +
  draw_grob(legend_guich, x = 0.88, y = 0.275, width = 0.01, height = 0.01)
#
ggsave("./output/figures/main/fig_mass.png", plot=fig_mass, width = 21, height = 8, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/main/fig_mass.png")
```

@@

_Acute stressor_: Latency to move after the first exposure in Control-Hot incubated _L. delicata_ was higher than for _L. guichenoti_ (I~L. deli~ - I~L. guich~ = `r format_dec(median(deli_move_posteriors$int_Control_Hot) - median(guich_move_posteriors$int_Control_Hot), 3)`; 95% HDPI = [`r format_dec(hdi(deli_move_posteriors$int_Control_Hot - guich_move_posteriors$int_Control_Hot, credMass = 0.95)[1], 3)`, `r format_dec(hdi(deli_move_posteriors$int_Control_Hot - guich_move_posteriors$int_Control_Hot, credMass = 0.95)[2], 3)`]), but there were no differences in Latency to move between the rest of the treatments (see @table_contrasts_species). In contrast, the probability of emerging from the shelter after an acute stressor was lower in cold incubated _L. delicata_ than _L. guichenoti_ for both control (I~L. deli~ - I~L. guich~ = `r format_dec(median(deli_emergence_posteriors$int_Control_Cold) - median(guich_emergence_posteriors$int_Control_Cold), 3)`; 95% HDPI = [`r format_dec(hdi(deli_emergence_posteriors$int_Control_Cold - guich_emergence_posteriors$int_Control_Cold, credMass = 0.95)[1], 3)`, `r format_dec(hdi(deli_emergence_posteriors$int_Control_Cold - guich_emergence_posteriors$int_Control_Cold, credMass = 0.95)[2], 3)`]) and CORT-treated (I~L. deli~ - I~L. guich~ = `r format_dec(median(deli_emergence_posteriors$int_CORT_Cold) - median(guich_emergence_posteriors$int_CORT_Cold), 3)`; 95% HDPI = [`r format_dec(hdi(deli_emergence_posteriors$int_CORT_Cold - guich_emergence_posteriors$int_CORT_Cold, credMass = 0.95)[1], 3)`, `r format_dec(hdi(deli_emergence_posteriors$int_CORT_Cold - guich_emergence_posteriors$int_CORT_Cold, credMass = 0.95)[2], 3)`]) lizards, but not in hot incubated animals (see @table_contrasts_species). There were no differences between species in the latency until sheltering after an acute stressor (see @table_contrasts_species).

>>>>

Animals constantly face events that can disrupt their homeostasis. These disruptions can be acute, such as a sudden predator attack, or long-term, like prolonged droughts or habitat degradation [@sapolsky_how_2000; @mcewen_concept_2003]. Regardless of the nature of the challenge, organisms must adjust to allostasis to maintain their physiological and behavioural integrity [@wingfield2002endocrine]. In vertebrates, the ability to cope with disruptions in homeostasis is carried out by the hypothalamic-pituitary-adrenal/interrenal (HPA/HPI) axis [@sapolsky_how_2000]. This system elicits a physiological response, also known as the stress response, mediated by glucocorticoids (GCs), a group of hormones that facilitate the reallocation of energetic resources to restore homeostasis [@sapolsky_how_2000; @mcewen_concept_2003]. For example, when an animal faces a predatory attack, GCs mobilise energy to support increased locomotor activity, vigilance, or attention to avoid the threat [@trompeter2011invader]. In the short term, GCs help organisms cope with acute challenges. However, long-term exposure to GCs can carry significant metabolic costs, leading to immune suppression, impaired growth, or permanent neural alterations [@picard_mitochondria_2014; @mcewen2017neurobiological]. Furthermore, reacting to stressors can reduce the time and energy available for essential activities such as foraging, thermoregulation, or social interactions [@belliure2004behavioral; @martin2024blind]. As such, animals are predicted to evolve mechanisms to appropriately respond to acute stressors while avoiding the costs in the long term. For example, by regulating their responses to homotypic (repeated and similar) stressors, particularly those that are predictable and non-lethal [@grissom2009habituation]. Habituation - the reduction of physiological responses elicited by exposure to a homotypic stressor [@pfister1979glucocorticosterone] - can be crucial to avoid the long-term costs of chronic stress without compromising the response to acute stressors.

Importantly, the process of physiological habituation shares key functional characteristics with the cognitive process of habituation - a form of non-associative learning where individuals reduce their responsiveness to a repeated stimulus [@thompson1966habituation]. In both cases, organisms are repeatedly exposed to a stimulus/homotypic stressor and progressively decrease their response toward it [@grissom2009habituation]. In fact, both behavioural and physiological habituation often co-occur. For example, rhesus macaques (_Macaca mulatta_) showed a decline in blood cortisol concentrations across consecutive days of restraint, accompanied by decreased vocalizations and behavioral agitation [@ruys2004behavioral]. How both types of habituation interplay with each other depends on the nature of the stressor. Physiological habituation does not occur when animals are faced with systemic stressors like hypoglycemia or ether exposure, whereas habituation to processive stressors — those involving primarily psychological qualities, such as restraint or simulated predation - is frequent [@grissom2009habituation]. However, physiological habituation alone is insufficient to explain the behavioural patterns observed when animals are repeatedly exposed to a homotypic processive stressor [@jaferi2006corticosterone; @grissom2009habituation]. For example, adrenalectomised rats - incapable of physiological habituation - still exhibit cognitive habituation [@jaferi2006corticosterone], suggesting a degree of independence between the two processes. While the HPA/HPI axis is paramount for physiological habituation, cognitive habituation also involves neural mechanisms related to sensory processing, learning, and behavioural control, such as the limbic system, the superior colliculus, or the optic tectum [@grissom2009habituation; @dutta2014saliency]. Habituation to processive stressors is therefore likely to involve complex interactions between physiological and cognitive processes.

Crucially, the effectiveness of either system depends on how well the underlying neural structures develop and integrate. Many of these brain regions mature concurrently and can be highly plastic during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017]. Early-life conditions can shape the structure and function of both neural systems, with long-term consequences for how individuals respond to acute and chronic stressors [@vargas2016early; @van2017modulation; @zito2017early; @crino_under_2020; @crino2024eggs]. One of the key mediators of early-life effects is GC exposure. Elevations in GCs during development exert potent effects on HPA/HPI axis function, altering the response to stressors later in life [@costantini2011meta; @vargas2016early; @zito2017early; @crino2024eggs]. For example, zebra finches (_Taeniopygia guttata_) nutritionally stressed during early-life had higher levels of CORT than controls at post-hatching day 25, while periodic maternal separation increased basal corticosterone levels in rats (_Rattus norvegicus_) [@vargas2016early; @zito2017early]. In parallel, developmental GC exposure can impact brain regions involved in learning, memory, and behavioural flexibility, affecting neural processes relevant to cognitive habituation [@lemaire_prenatal_2000; @zhu_prenatal_2004; @costantini2011meta]. Therefore, early-life exposure to GCs can have long-lasting effects on the response to both acute and chronic stressors, potentially leading to maladaptive responses in adulthood.

@@

In this study, we investigate how prenatal CORT and incubation temperature shape the response to a stressor in two species of skinks: the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We focus on the effects of these early-life conditions on the behavioural responses to a processive stressor. Specifically, we examine how these factors influence the acute (i.e., the first exposure) and the chronic response (i.e., repeated exposures over time) responses to a simulated predatory attack. Finally, we assessed body mass change across the stress period as a potential indirect cost, while controlling for food intake. We predicted that prenatal CORT exposure would heighten acute stress reactivity and slow habituation, while warmer incubation temperatures would reduce acute responses and enhance habituation. Furthermore, we predicted higher temperatures would buffer the effects of prenatal CORT on stress responses, potentially leading to more flexible behavioural adjustment. Finally, we expected that less flexible stress responses would lead to higher decreases in mass.

@@

We used two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). These are oviparous, generalist, small (∼35-55 mm snout-vent length (SVL)) lizards that are sympatric in suburban areas throughout south-eastern Australia [@chapple_know_2011; @chapple_biology_2014]. However, _L. delicata_ has been reported as a successful invader in New Zealand and Hawaii, while there are no such reports for _L. guichenoti_ [@chapple_know_2011; @chapple_biology_2014]. Invassiveness does not seem to be related to differences in cognitive abilities [@bezzina2014does; @recio2025early; @recio2025early], but seems to be associated with differences in personality [@chapple_know_2011], which may be linked to differences in their ability to respond to acute and chronic processive stressors [@koolhaas1999coping].

@@

*Breeding colony* - Lizards came from a breeding colony established in the laboratory since 2019. This colony consisted of 270 adults of _L. delicata_ and 180 adults of _L. guichenoti_ housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with two males and four females per enclosure. Enclosures were provided with shelter, nonstick matting, and several small water dishes. The lizards were given water daily and were fed approximately 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. The crickets were dusted with calcium weekly and multivitamins and calcium biweekly. Room temperature was set to 22-24 ºC, but we also provided the enclosures with a heat cord and a heat lamp following a 12 h light:12 h dark cycle, keeping the warm side of the enclosures at 34 ºC.

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them at cold (23 ± 3 ºC) or hot (28 ± 3 ºC) conditions (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study, where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Mean incubation temperatures represent the lower and upper limits of the natural range of nest temperatures in _L. delicata_ [mean nesting temperatures = 27.4 ºC; @cheetham2011embryonic]. Nonetheless, the higher temperature treatment (28 ºC) was above the thermal optima estimated for _L. delicata_ [T~opt~ = 25 ºC; @pettersen2023maternal].

@@

To test the effects of prenatal conditions on the response to a stressor, we simulated a predatory attack - a standardized processive stressor [@thaker2009acute; @trompeter2011invader] - once daily over multiple days. Each trial began with one of the researchers (PR) removing the water dish from the enclosure, followed by the shelter. The lizard was then chased for 60 seconds using a soft paintbrush, simulating the predatory attack. Immediately after the chase, the shelter — but not the water dish — was returned to its original position, marking the start of the behavioural observation period. The lizard’s behaviour was then recorded for one hour. All videos were later analysed by MD, who was blinded to the lizards’ treatment. From each video, we recorded three behavioural variables: i) latency to move — the time spent immobile immediately after the simulated attack; ii) latency to shelter — the time it took the lizard to hide under the shelter after resuming movement; and iii) emergence — whether the lizard emerged from the shelter during the 40 minutes following hiding (1 = emerged, 0 = did not emerge). A lizard was considered to have resumed movement if it moved continuously for at least five seconds, and to be under shelter if no limbs were visible. This procedure was repeated daily for eight consecutive days to assess the effects of long-term exposure to a processive stressor. All trials were conducted between 1000–1400 h, when lizards were most active.

Both before and after the completion of the behavioural tests, lizard mass was measured to the nearest 0.001 g using a digital scale. Change in mass was calculated as the difference between the final minus the initial mass. Change in mass was used as a proxy for the metabolic cost of repeated stress exposure, where mass loss or lower growth was considered as higher costs. To control for hunger levels and potential effects of food ingestion on mass change, we also recorded the number of crickets ingested by each lizard during the acclimation period. To do so, we placed a known number of crickets in each enclosure after each of the trials, and counted the number of crickets remaining after 24 hours. The total number of crickets ingested during the eight days of tests was then used as a covariate in the analyses of mass change. The experimental design is summarised in @fig-Methods B.

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Scheme of our experimental design. In panel A, the early environment manipulations leading to four experimental conditions. In panel B, the timeline of the postnatal stress-response assay: on Day 0 and 9, lizard mass was recorded; from Days 1–8, lizards were subjected daily to a simulated predator attack (chased with a soft paintbrush for 1 minute), followed by a 1-hour behavioural recording. Variables measured the latency to move, the latency to shelter, and the probability of emergening from the shelter in 40 minutes after the stressor."

@@

We performed the analyses for each species separately. We fitted a total of eight models using the _brm_ package [@burkner2017brms], which fits Bayesian multilevel models with _Stan_ [@stan] with R version 4.4.0 [@R]. We ran four parallel MCMC chains of 10000 iterations for each model, with a warmup period of 4000 iterations.

We used as the response variables the latency to move, the latency to shelter, and the emergence probability, as well as the change in mass after the eight days of stress exposure. We fitted separate models for each response variable. For the three behavioural variables, we included the prenatal condition (Control-Cold, CORT-Cold, Control-Hot, CORT-Hot), the day of the trial, and their interaction as fixed effects. To facilitate interpretation of the model intercepts, trial day was transformed so that day 1 in the experiment corresponds to day 0 in the model (i.e., the intercept reflects lizard behaviour in response to the first exposure to the stressor). For the variable mass, we included the prenatal condition and the total number of crickets ingested per each individual during the eight days of stress exposure as predictors.

We included each lizard's random intercept and slope (trial) as a random factor in the behavioural models, but not when mass was the response. For all models, we included the clutch as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple2015deliinvLHI]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

@@

All the estimated parameters in the results were transformed back to original units employing appropriate formulas depending on the distribution employed in the models. The error structure for latency to move, latency to shelter, or mass change followed a lognormal distribution [family = lognormal()], while the probability of emerging from the shelter was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = 'logit')]. Change in mass was also rescaled by adding the minimum change in mass recorded before running the models to avoid negative values but transformed back to original values for the presentation of the results.

@@

_Lampropholis delicata_: Incubation at cold temperatures decreased the latency to move after a simulated attack on the first day (Intercept (denoted hereafter as I)~Hot~ - I~Cold~ = `r deli_move_contrasts$median_Temperature_int`; 95% Highest Posterior Density Intervals (95% HPDI) = [`r deli_move_contrasts$X2.5.HDI_Temperature_int`, `r deli_move_contrasts$X97.5.HDI_Temperature_int`]), while neither CORT or the interaction have any effect on latency to shelter or emergence (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_). We did not find any effects of prenatal temperature, CORT or their interaction on latency to shelter or emergence probability after an acute stressor (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_).

@@

```{r, fig-results_mass}
#| label: fig-results_mass
#| fig.cap: "Change in mass of _Lampropholis delicata_ and _Lampropholis guichenoti_ after eight days of repeated stress exposure." 
#
#
source(here("R", "func.R"))
#
## L delicata
deli_mass_plot <- plotting(sp = "deli", res = "mass", type = "int") +
  theme(legend.position = "none",
        legend.title = element_blank())
# Getting the legend
legend_plot_deli <- plotting(sp = "deli", res = "mass", type = "int") + 
  theme(legend.position = "right", legend.title = element_blank())
gtable_deli <- ggplotGrob(legend_plot_deli)
legend_deli <- gtable_deli$grobs[[which(sapply(gtable_deli$grobs, function(x) x$name) == "guide-box")]]
#
## L guichenoti
guich_mass_plot <- plotting(sp = "guich", res = "mass", type = "int") +
  theme(legend.position = "none",
        legend.title = element_blank())
# Getting the legend
legend_plot_guich <- plotting(sp = "guich", res = "mass", type = "int") + 
  theme(legend.position = "right", legend.title = element_blank())
gtable_guich <- ggplotGrob(legend_plot_guich)
legend_guich <- gtable_guich$grobs[[which(sapply(gtable_guich$grobs, function(x) x$name) == "guide-box")]]
#
mass_grid <- plot_grid(deli_mass_plot, NULL, guich_mass_plot, NULL,
                       ncol = 4, nrow = 1,
                       rel_widths = c(0.5, 0.5, 0.5, 0.45))
#
fig_mass <- ggdraw() +
  # Make background white
  theme(plot.background = element_rect(fill = "white", color = NA)) +
  # Add the final plot but with reduced size for the legend and "title"
  draw_plot(mass_grid, x = 0, y = 0, width = 1, height = 0.92) +
  # Insert species name
  annotate("text", x = 0.36, y = 0.99, label = "Lampropholis delicata", hjust = 1, vjust = 1, size = 5, fontface = "italic") +
  annotate("text", x = 0.9, y = 0.99, label = "Lampropholis guichenoti", hjust = 1, vjust = 1, size = 5, fontface = "italic") +
  # Insert title for each plot
  annotate("text", x = 0.023, y = 0.91, label = "A", hjust = 1, vjust = 1, size = 6, fontface = "bold") +
  annotate("text", x = 0.535, y = 0.91, label = "B", hjust = 1, vjust = 1, size = 6, fontface = "bold") +
  # Insert picture species
  draw_image(here("./Others/Deli.png"), x = 0.271, y = 0.65, width = 0.22, height = 0.22) +
  draw_image(here("./Others/Guich.png"), x = 0.78, y = 0.65, width = 0.22, height = 0.22) +
  # Insert legends
  draw_grob(legend_deli, x = 0.375, y = 0.275, width = 0.01, height = 0.01) +
  draw_grob(legend_guich, x = 0.88, y = 0.275, width = 0.01, height = 0.01)
#
ggsave("./output/figures/main/fig_mass.png", plot=fig_mass, width = 21, height = 8, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/main/fig_mass.png")
```

@@

```{r, contrasts_sp}
#| label: contrasts_sp
# Chunk created to make the df for the table below (Table 1) and also facilitate the comparisons between species below
#
contrasts_sp <- bind_rows(move_deli = deli_move_posteriors,
                          move_guich = guich_move_posteriors,
                          shelter_deli = deli_shelter_posteriors,
                          shelter_guich = guich_shelter_posteriors,
                          emergence_deli = deli_emergence_posteriors,
                          emergence_guich = guich_emergence_posteriors,
                          mass_deli = deli_mass_posteriors,
                          mass_guich = guich_mass_posteriors,
                          .id = "data") %>%
                pivot_longer(cols = -c(data),
                            names_to = c("stress", "treatment"),
                            names_pattern = "(int|slope)_(.*)") %>%
                # Pivot to wide format
                mutate(column_id = paste(data, stress, treatment, sep = "_")) %>%
                group_by(column_id) %>%
                mutate(draw = row_number()) %>%
                ungroup() %>%
                dplyr::select(column_id, draw, value) %>%
                pivot_wider(names_from = column_id, values_from = value) %>%

                #Get the real values
                mutate(across(matches("emergence.*_int"), ~ plogis(.x)),
                      across(matches("(move|shelter).*_int"), ~ exp(.x)),
                      across(matches("mass.*_int"), ~ exp(.x) + min(mass_clean$delta_mass, na.rm = TRUE) - 1)) %>%
data.frame()
```
